---
title: "01-find-matched-biospecimen"
author: "Aditya Lahiri, Eric Wafula, Jo Lynne Rokita"
date: "10/13/2022"
output: html_notebook
---
#Load Libraries

```{r load libraries}
suppressPackageStartupMessages({
library(dplyr)
library(data.table)
library(tidyverse)
library(ggplot2)
})
```

#Set up directories
```{r setup directories}
# root directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# data directory
dataDir <- file.path(root_dir, "data")

# module directory
moduleDir <- file.path(root_dir, "analyses","molecular-subtyping-NBL")

#results directory
results_dir <- "results"
# create if doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

#plot directory
plot_dir <- "plots"
# create if doesn't exist
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir)
}
```

Code logic: 
MYCN_DF has 1508 records, some of the IDs in Kids_First_Biospecimen_ID
are DNA IDs while others are RNA IDs. Among the DNA IDs, some have corresponding 
RNA IDs, so we need to first find the matching DNA/RNA IDs.  To find matching RNA 
IDs for DNA IDs, we create a match ID by combining Kids_First_Participant_ID + Sample_ID + Cohort. 
To this end, we create a new column in MYCN_Df called match_id. 

# Read the mycn-nbl file 
```{r read the file containing mycn-nbl biospecimen}
MYCN_DF<-readr::read_tsv(file.path(results_dir, "mycn_nbl_subset_data.tsv"),guess_max = 100000)
```


# Create dataframe of samples where DNA and RNA IDs match 
```{r create match ID and dataframes of RNA and DNA}
# Create Match_ID Kids_First_Participant_ID + Sample_ID + Cohort
MYCN_DF <- MYCN_DF %>%mutate(match_id = paste(Kids_First_Participant_ID, 
                                              sample_id,cohort,sep = "_"))

# DNA samples are those belonging to experimental_strategy: WGS, WXS, and Targeted Sequencing  
# Some targeted sequencing also have RNA_Library as "exome capture" those biospecimens are RNAs.

MYCN_dna_df<-MYCN_DF %>% filter(is.na(RNA_library),experimental_strategy %in% 
                                  c("WGS", "WXS", "Targeted Sequencing")) # 1103 samples samples


MYCN_rna_df<-MYCN_DF %>% filter( (experimental_strategy=="RNA-Seq") | 
                                   (experimental_strategy=="Targeted Sequencing" & 
                                      !is.na(RNA_library)))
# 405 samples 374 rna-seq + 31 targeted_seq+exome_capture

# Rename the columns
colnames(MYCN_dna_df)[which(names(MYCN_dna_df)
                            == "Kids_First_Biospecimen_ID")] <- "Kids_First_Biospecimen_ID_DNA"
colnames(MYCN_rna_df)[which(names(MYCN_rna_df)
                            == "Kids_First_Biospecimen_ID")] <- "Kids_First_Biospecimen_ID_RNA"
```

Code Logic:
Now find the intersection of MYCN_dna_df and MYCN_rna_df, this will give us a dataframe 
of matching DNA and RNA biospecimen IDs. We then select specific columns from the joined dataframe 
for downstream analysis. 

```{r Further filter the data to create alteration table}
MYCN_DNA_RNA_DF <- inner_join(MYCN_dna_df,MYCN_rna_df, by="match_id")  # 342 samples 

 # Retain the information of RNA/DNA Biospecimen ID, gene symbol, cancer group, 
 # pathology free text diagnosis, copy number, TPM and status
MYCN_DNA_RNA_FILTERED_DF <- MYCN_DNA_RNA_DF %>% select(Kids_First_Biospecimen_ID_DNA,
                                                       Kids_First_Biospecimen_ID_RNA,
                                                       Kids_First_Participant_ID.x,
                                                       Kids_First_Participant_ID.y,
                                                       gene_symbol.x,gene_symbol.y,
                                                       pathology_diagnosis.x,pathology_diagnosis.y, 
                                                       pathology_free_text_diagnosis.x,
                                                       pathology_free_text_diagnosis.y, 
                                                       copy_number.x,copy_number.y,MYCN_TPM.x,
                                                       MYCN_TPM.y,status.x,status.y, match_id)
# Remove columns with no info, the following code logic is adapted from [2] 
empty_columns <- colSums(is.na(MYCN_DNA_RNA_FILTERED_DF) 
                         | MYCN_DNA_RNA_FILTERED_DF == "") == nrow(MYCN_DNA_RNA_FILTERED_DF)

#Remove empty coulmns
MYCN_DNA_RNA_FILTERED_DF<- MYCN_DNA_RNA_FILTERED_DF[!empty_columns]

#Check if the copies of pathology_diagnosis, pathology_free_text_diagnosism status are identitical. 
#These copies are created due to inner join of MYCN_dna_df and MYCN_rna_df
print(all.equal(MYCN_DNA_RNA_FILTERED_DF$pathology_diagnosis.x,
                MYCN_DNA_RNA_FILTERED_DF$pathology_diagnosis.y))

print(all.equal(MYCN_DNA_RNA_FILTERED_DF$pathology_free_text_diagnosis.x,
                MYCN_DNA_RNA_FILTERED_DF$pathology_free_text_diagnosis.y))

print(all.equal(MYCN_DNA_RNA_FILTERED_DF$Kids_First_Participant_ID.x,
                MYCN_DNA_RNA_FILTERED_DF$Kids_First_Participant_ID.y))
# since they are identical we can remove the duplicated columns.

# Remove duplicate columns after inspection
duplicate_columns <- c("pathology_diagnosis.y","pathology_free_text_diagnosis.y",
                       "Kids_First_Participant_ID.y")
MYCN_DNA_RNA_FILTERED_DF<-MYCN_DNA_RNA_FILTERED_DF[ , -which(names(MYCN_DNA_RNA_FILTERED_DF)
                                                             %in% duplicate_columns)] # 342 samples 


# Rename the columns appropriately.
colnames(MYCN_DNA_RNA_FILTERED_DF)<-c("Kids_First_Biospecimen_ID_DNA","Kids_First_Biospecimen_ID_RNA",
                                      "Kids_First_Participant_ID","gene_symbol","cancer_group",
                                      "pathology_free_text_diagnosis","copy_number","MYCN_TPM",
                                      "status","Match_ID")    
```


Code Logic:
In MYCN_DNA_RNA_FILTERED_DF we want to find the repeating Kids_First_Biospecimen_ID_DNA 
We do not need to check for repeating Kids_First_Biospecimen_ID_RNA as all the DNA and RNA IDs are 
matched in MYCN_DNA_RNA_FILTERED_DF. Therefore if we find a repeating DNA ID, we will also find its 
corresponding RNA ID which will also be repeating. 

```{r check for repeating Kids_First_Biospecimen_ID_DNA}
# We will check only repeating DNA IDs as the records unique in MYCN_DNA_RNA_FILTERED_DF 
# when we query them with DNA IDs + RNA IDs. 
MYCN_Table <- as.data.frame(table(MYCN_DNA_RNA_FILTERED_DF$Kids_First_Biospecimen_ID_DNA))
colnames(MYCN_Table)<- c("Kids_First_Biospecimen_ID_DNA", "Freq")
MYCN_Table$Kids_First_Biospecimen_ID_DNA<-as.character(MYCN_Table$Kids_First_Biospecimen_ID_DNA)
index_repeats <- which(MYCN_Table$Freq>1)
repeated_ids <- MYCN_Table$Kids_First_Biospecimen_ID_DNA[index_repeats]
for (iter in 1:length(repeated_ids))
  print(MYCN_DNA_RNA_FILTERED_DF[MYCN_DNA_RNA_FILTERED_DF$Kids_First_Biospecimen_ID_DNA
                                 ==repeated_ids[iter],])
```



Code Logic:
We find that there are four DNA IDs which are repeating. 
For the record TARGET-30-PAPVFD-01A-01D retaining the record with status Neutral and copynumber 3
For the record TARGET-30-PASXRG-01A-01D retaining the record with copy number 14
For the records BS_0XC02E11 and BS_1F8J25Q1 we see that in their repeating copies all column values 
except for their aliquot_id are different. We remove the rows with aliquot_ids of 
ET_FD9T78QE_DGD_FUSIP_29 for the time being, in each sample. 

```{r removing values}
# For records BS_0XC02E11 and BS_1F8J25Q1 two samples are there, differing in aliquot_id.y
rows_to_delete_MYCN <- c(which(MYCN_DNA_RNA_FILTERED_DF$Kids_First_Biospecimen_ID_DNA
                               =="TARGET-30-PAPVFD-01A-01D" & 
                                 MYCN_DNA_RNA_FILTERED_DF$copy_number==2),
                        which(MYCN_DNA_RNA_FILTERED_DF$Kids_First_Biospecimen_ID_DNA
                              =="TARGET-30-PASXRG-01A-01D" & MYCN_DNA_RNA_FILTERED_DF$copy_number==4))

rows_to_delete_MYCN_DGD_1<-which(MYCN_DNA_RNA_FILTERED_DF$Kids_First_Biospecimen_ID_DNA
                                 =="BS_0XC02E11") [2] # remove ET_FD9T78QE_DGD_FUSIP_29

rows_to_delete_MYCN_DGD_2<-which(MYCN_DNA_RNA_FILTERED_DF$Kids_First_Biospecimen_ID_DNA
                                 =="BS_1F8J25Q1") [2] # ET_FD9T78QE_DGD_FUSIP_29 

# 338 samples 342- 4 (repeating samples) = 338
MYCN_DNA_RNA_FILTERED_DF<- MYCN_DNA_RNA_FILTERED_DF[-c(rows_to_delete_MYCN,
                                                       rows_to_delete_MYCN_DGD_1,
                                                       rows_to_delete_MYCN_DGD_2),] 
```

Code Logic:
Plot TPMs vs Biospecimen for samples that have both DNA and RNA IDs. Only plot those biospecimens 
that have a TPM value. 
```{r make TPM vs Biospecimen plot for samples with matched DNA and RNA IDs}
#Select relevant info
MYCN_Plot_df <- MYCN_DNA_RNA_FILTERED_DF %>% filter(!is.na(MYCN_TPM)) %>%
  select (Kids_First_Biospecimen_ID_DNA,MYCN_TPM,status) 

# Refered code from [5]
bar_plot1<-MYCN_Plot_df %>% 
  arrange(MYCN_TPM) %>% 
  mutate(Kids_First_Biospecimen_ID_DNA = factor(Kids_First_Biospecimen_ID_DNA, 
                                                unique(Kids_First_Biospecimen_ID_DNA))) %>%
  ggplot() +aes(x=Kids_First_Biospecimen_ID_DNA, y=MYCN_TPM, fill=status)+
  geom_bar(position="dodge",stat="identity")+
  theme(axis.text.x = element_text(size = 5, angle = 45, hjust = 1))


plot_file <- file.path(plot_dir, "TPM_Biospecimen_Matching.png")
ggsave(filename = plot_file,width = 45, height = 20)
```


# Write the table containing mycn-nbl biospecimen with matched DNA and RNA Ids into results dir
```{r write alteration table}
MYCN_DNA_RNA_FILTERED_DF%>%readr::write_tsv(file.path(results_dir, "mycn_nbl_matched_biospecimen.tsv"))
```
