---
title: "01-find-matched-biospecimen"
author: "Aditya Lahiri, Eric Wafula, Jo Lynne Rokita"
date: "10/13/2022"
output: html_notebook
---
In this notebook we load the table `nbl-subset/mycn_nbl_subset_data.tsv` and find the
biospecimen which have matched DNA and RNA IDs. We store these biospecimen as a table 
in `nbl-subset/mycn_nbl_matched_biospecimen.tsv`. The script also creates two tables,
`nbl-subset/mycn_nbl_dna_biospecimen.tsv` and `nbl-subset/mycn_nbl_rna_biospecimen.tsv`
which contain biospecimen whose `Kids_First_Biospecimen_ID` are DNA and RNA respectively. 

The notebook also plots the `TPM` vs `Kids_First_Biospecimen_ID` for biospecimen with
matched DNA and RNA IDs. We overlay the consensus status call for each of the biospecimen 
in this plot.

## Load Libraries

```{r load libraries}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggplot2))
```

## Set up directories

```{r setup directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
module_dir <- file.path(root_dir, "analyses","molecular-subtyping-NBL")
subset_dir <- file.path(module_dir, "nbl-subset")
results_dir <- file.path(module_dir, "results")
plot_dir <- file.path(module_dir, "plots")

# create plot_dir if doesn't exist
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir)
}
```

Create a `match_id` combining `Kids_First_Participant_ID`, `sample_id`, and
`cohort` for determining matching DNA-RNA bioscpecimen.

## Load MYCN subset file 

```{r mycn subset file}
mycn_df <- readr::read_tsv(file.path(subset_dir, "mycn_nbl_subset_data.tsv"), 
                           guess_max = 100000) %>% 
  # create match_id
  dplyr::mutate(match_id = paste(Kids_First_Participant_ID, 
                                 sample_id, cohort, sep = "_")) %>%
  # select columns of interest
  dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, aliquot_id, 
                experimental_strategy, RNA_library,  pathology_diagnosis, cohort,
                pathology_free_text_diagnosis, gene_symbol, MYCN_CN, MYCN_TPM,
                MYCN_CN_status, match_id)
```

## DNA and RNA biospecimen

```{r dna and rna biospecimen}
# DNA biospecimen
mycn_dna_df <- mycn_df %>% 
  dplyr::filter(experimental_strategy %in% 
                  c("WGS", "WXS", "Targeted Sequencing"),
                is.na(RNA_library)) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID)

# RNA biospecimen
mycn_rna_df <- mycn_df %>%
  dplyr::filter(experimental_strategy %in% c("RNA-Seq","Targeted Sequencing"),
                !is.na(RNA_library)) %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID)

```

## Determine matching DNA-RNA biospecimen

```{r matching dna and rna}
mycn_dna_rna_df <- mycn_dna_df %>% 
  dplyr::select(Kids_First_Participant_ID, gene_symbol, pathology_diagnosis,
                pathology_free_text_diagnosis, MYCN_CN, MYCN_CN_status,
                Kids_First_Biospecimen_ID_DNA, match_id, aliquot_id, cohort) %>% 
  dplyr::inner_join(
    mycn_rna_df %>% 
      dplyr::select(Kids_First_Biospecimen_ID_RNA, MYCN_TPM, match_id),
    by = "match_id")
  
```

Determine of duplicate DNA (`Kids_First_Biospecimen_ID_DNA`) and RNA 
biospecimen (`Kids_First_Biospecimen_ID_RNA`) in the matched DNA-RNA dataframe. 
We do not need to check for duplicate RNA as all the DNA and RNA biospecimen are
matched. Therefore, if we find a duplicate DNA biospecimen, we will also find its 
corresponding RNA biospecimen that is also a duplicate. 

## Find duplicate DNA biospecimen

```{r duplicate biospecimen}
# get duplicate dna IDs
dup_dna_ids <- mycn_dna_rna_df %>% 
  count(Kids_First_Biospecimen_ID_DNA) %>% 
  dplyr::filter(n > 1) %>% 
  pull(Kids_First_Biospecimen_ID_DNA)

# display duplicate biospecimen
mycn_dna_rna_df %>% dplyr::filter(Kids_First_Biospecimen_ID_DNA %in% 
                                    dup_dna_ids) 

```

There are four pairs of duplicate DNA and RNA biospecimen. 
Biospecimen with lower copy number in each duplicate pair will be left out. 
`Adyita` and `Jo Lynne` determine the two `DGD` bioscpecimen to exclude. 

## Remove dupicate biospecimen

```{r remove duplicates}
# matched biospecimen
mycn_dna_rna_df <- mycn_dna_rna_df %>% 
  dplyr::filter(!(Kids_First_Biospecimen_ID_DNA == "TARGET-30-PAPVFD-01A-01D" & 
                   MYCN_CN == 2), 
                !(Kids_First_Biospecimen_ID_DNA == "TARGET-30-PASXRG-01A-01D" &
                   MYCN_CN == 4),
                !(Kids_First_Biospecimen_ID_DNA == "BS_1F8J25Q1" &
                   Kids_First_Biospecimen_ID_RNA == "BS_2HM4AE24"),
                !(Kids_First_Biospecimen_ID_DNA == "BS_0XC02E11" &
                   Kids_First_Biospecimen_ID_RNA == "BS_XV40PE3Q"))

```

Plot `TPM vs Biospecimen` for RNA biospecimen that are matched with DNA. 
Only plot those biospecimens that have a TPM value.

## Plot distribution of MYCN TPM values

```{r plot tpm distribution}
# get relevant variables
mycn_plot_df <- mycn_dna_rna_df %>% 
  dplyr::filter(!is.na(MYCN_TPM)) %>%
  dplyr::select(Kids_First_Biospecimen_ID_DNA, MYCN_TPM, MYCN_CN_status) 

# plot distribution
bar_plot1 <- mycn_plot_df %>% 
  dplyr::arrange(MYCN_TPM) %>% 
  dplyr::mutate(Kids_First_Biospecimen_ID_DNA = 
                  factor(Kids_First_Biospecimen_ID_DNA, 
                         unique(Kids_First_Biospecimen_ID_DNA))) %>%
  ggplot() + aes(x = Kids_First_Biospecimen_ID_DNA, y = MYCN_TPM, 
                 fill = MYCN_CN_status) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(size = 5, angle = 45, hjust = 1))

# save distribution plot to file
plot_file <- file.path(plot_dir, "tpm_biospecimen_matching.png")
ggsave(filename = plot_file, width = 45, height = 20)
```

Write the table containing mycn-nbl biospecimen with matched DNA and RNA 
biospecimen, DNA only, and RNA only to file

## Write tables to file

```{r write alteration table}
mycn_dna_rna_df %>% 
  readr::write_tsv(file.path(subset_dir,"mycn_nbl_matched_biospecimen.tsv"))
mycn_dna_df %>% 
  readr::write_tsv(file.path(subset_dir,"mycn_nbl_dna_biospecimen.tsv"))
mycn_rna_df %>% 
  readr::write_tsv(file.path(subset_dir,"mycn_nbl_rna_biospecimen.tsv"))

```

## Session Info

```{r sessioninfo}
sessionInfo() 
```
