---
title: "02-find-non-matching-biospecimen"
output: html_notebook
author: "Aditya Lahiri, Eric Wafula, Jo Lynne Rokita"
date: "10/13/2022"
---

In this notebook we find the biospecimen which do not have matched DNA and RNA 
biospecimen. We create a dataframe for these biospecimens called `MYCN_non_match_df` 
which has all the necessary information required for performing subtyping. We 
concatenate this `MYCN_non_match_df` with the table `nbl-subset/mycn_nbl_matched_biospecimen.tsv` 
and joined table called `nbl-subset/alteration_table_without_subtype.tsv` which 
contains the list of all NBL-MYCN biospecimen that need to subtpyped. 

The notebook also plots the `TPM vs Kids_First_Biospecimen_ID` for all the biospecimen 
for which there is a TPM value. We overlay the consensus status call for each of 
the biospecimen in this plot and determine a cut off for amplification. 


```{r load libraries}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
```

## Set up directories
```{r setup directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
module_dir <- file.path(root_dir, "analyses","molecular-subtyping-NBL")
subset_dir <- file.path(module_dir, "nbl-subset")
results_dir <- file.path(module_dir, "results")
plot_dir <- file.path(module_dir, "plots")

```

## Load files

```{r load files}
# mycn subset and filtered file
mycn_df <- 
  readr::read_tsv(file.path(subset_dir, "mycn_nbl_subset_data.tsv"), 
                  guess_max = 10000)
mycn_dna_rna_df <- 
  readr::read_tsv(file.path(subset_dir, "mycn_nbl_matched_biospecimen.tsv"), 
                  guess_max = 10000)

# gmkf clinical patient mycn amplification status
gmkf_mycn_status_df <- 
  readr::read_tsv(file.path(subset_dir, 
                            "gmkf_patient_clinical_mycn_status.tsv"))

target_mycn_status_df <- 
  readr::read_tsv(file.path(subset_dir, 
                            "target_patient_clinical_mycn_status.tsv")) 

```

## Unmatched DNA and RNA biospecimen

```{r unmatched biospecimen}
# unmatched dna biospecimen
mycn_dna_df <- mycn_df %>% 
  dplyr::filter(experimental_strategy %in% 
                  c("WGS", "WXS", "Targeted Sequencing"),
                is.na(RNA_library), 
                !(Kids_First_Biospecimen_ID %in%
                  mycn_dna_rna_df$Kids_First_Biospecimen_ID_DNA)) %>% 
  dplyr::rename(Kids_First_Biospecimen_ID_DNA = Kids_First_Biospecimen_ID) %>% 
  dplyr::select(Kids_First_Biospecimen_ID_DNA, Kids_First_Participant_ID, MYCN_TPM,
                MYCN_CN, MYCN_CN_status, pathology_free_text_diagnosis, cohort) %>% 
  dplyr::mutate(Kids_First_Biospecimen_ID_RNA = NA_character_)
  
# unmatched rna biospecimen
mycn_rna_df <- mycn_df %>%
  dplyr::filter(experimental_strategy %in% c("RNA-Seq","Targeted Sequencing"),
                !is.na(RNA_library),
                !(Kids_First_Biospecimen_ID %in%
                  mycn_dna_rna_df$Kids_First_Biospecimen_ID_RNA)) %>%
  dplyr::rename(Kids_First_Biospecimen_ID_RNA = Kids_First_Biospecimen_ID) %>% 
  dplyr::select(Kids_First_Biospecimen_ID_RNA, Kids_First_Participant_ID, MYCN_TPM,
                MYCN_CN, MYCN_CN_status, pathology_free_text_diagnosis, cohort) %>% 
  dplyr::mutate(Kids_First_Biospecimen_ID_DNA = NA_character_)

```

## Merge matched and unmatched biospecimen

```{r merge biospecimen}
# select relevant columns from matched dna-rna biospecimen dataframe
mycn_dna_rna_df <- mycn_dna_rna_df %>% 
  dplyr::select(Kids_First_Biospecimen_ID_DNA, Kids_First_Biospecimen_ID_RNA,
                Kids_First_Participant_ID, MYCN_TPM, MYCN_CN ,MYCN_CN_status,
                pathology_free_text_diagnosis, cohort) 

# bind matched and unmatched biospecimen 
alteration_df <- dplyr::bind_rows(mycn_dna_rna_df, mycn_dna_df, mycn_rna_df)

```

## Add clinical MYCN status

```{r add clinical mycn status}
# add gmkf clinical mycn status
gmkf_mycn_df <- alteration_df %>% 
  dplyr::filter(cohort == "GMKF") %>% 
  dplyr::select(-pathology_free_text_diagnosis) %>% 
  dplyr::left_join(gmkf_mycn_status_df, 
                   by = c("Kids_First_Participant_ID" = "USI_ID")) %>% 
  dplyr::rename(pathology_free_text_diagnosis =
                  GMKF_MYCN_STATUS)

# add target clinical mycn status
target_mycn_df <- alteration_df %>% 
  dplyr::filter(cohort == "TARGET") %>% 
  dplyr::select(-pathology_free_text_diagnosis) %>% 
  dplyr::left_join(target_mycn_status_df, 
                   by = c("Kids_First_Participant_ID" = "USI_ID")) %>% 
  dplyr::rename(pathology_free_text_diagnosis =
                  TARGET_MYCN_STATUS)

# other cohorts
others_mycn_df <- alteration_df %>% 
  dplyr::filter(!cohort %in% c("GMKF", "TARGET"))

# binc cohort biospecimen 
alteration_df <- dplyr::bind_rows(gmkf_mycn_df, target_mycn_df, others_mycn_df)

```

Plot the TPM values for biospecimens which have TPMs both for  Matched and Non Matched  
biospecimens.

## Determine MYCN expression cutoff 

```{r plot tpm distribution}
plot_alteration_df <- alteration_df %>% 
  dplyr::filter(!is.na(MYCN_TPM)) %>% 
  dplyr::mutate(Kids_First_Biospecimen_ID = 
                  coalesce(Kids_First_Biospecimen_ID_DNA,
                           Kids_First_Biospecimen_ID_RNA))

bar_plot2<-plot_alteration_df %>%
  dplyr::arrange(MYCN_TPM) %>%
  dplyr::mutate(Kids_First_Biospecimen_ID = 
                  factor(Kids_First_Biospecimen_ID,
                         unique(Kids_First_Biospecimen_ID))) %>%
  ggplot() + aes(x = Kids_First_Biospecimen_ID, 
                 y = MYCN_TPM, 
                 fill = MYCN_CN_status) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(size = 2, angle = 45, hjust = 1))
# display plot
bar_plot2

Suggested_Cutoff <- 140.83
bar_plot2 + geom_hline(aes(yintercept=Suggested_Cutoff, 
                           linetype = "Suggested Amp Cutoff"),
                       color = "black", size= 2)

plot_file <- file.path(plot_dir, "tpm_biospecimen_all_samples_with_tpm.png")
ggsave(filename = plot_file, width = 49, height = 20)
```


## Write to file MYCN biospecimen to subtype

```{r write output to file}
alteration_df %>% 
  readr::write_tsv(file.path(subset_dir, 
                             "alteration_table_without_molecular_subtype.tsv"))

```

## Session Info

```{r sessioninfo}
sessionInfo() 
```
