---
title: "03-subtyping"
author: "Aditya Lahiri, Eric Wafula, Jo Lynne Rokita"
date: "11/14/2022"
output: html_notebook
---
## Objective
To subtype `Neuroblastoma`, `Ganglioneuroblastoma`, and `Ganglioneuroma` biospecimen into either `MYCN amplified` or `MYCN non-amplified`. 

This script loads the table `input/alteration_table_without_molecular_subtype.tsv`and determines MYCN molecular subtypes based on the six subtyping criteria described in module README. This script generates a table called `results/neuroblastoma_molecular_subtypes.tsv`, which contains the molecular subtypes for each of the biospecimen in the table `alteration_df`.

This script also creates the following additional result files:

1. `results/alteration_table_with_molecular_subtype.tsv` files contains all the biospecimen for Neuroblastama, Ganglioneuroblastoma, and Ganglioneuroma with more detailed descriptive MYCN molecular subtyping annotations. In contrast, `results/neuroblastoma_molecular_subtypes.tsv` files summarized to exclude detailed descriptive information why a biospecimen was or wasn't subtyped and suitable for usage for subsequent analyses. 

2. `molecular_subtypes_Based_On_Cutoff.tsv` contains the a lists of biospecimen that were subtyped based on a gene expression TPM thresholding (cutoff).

 

## Load Libraries

```{r load libraries}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggbio))
suppressPackageStartupMessages(library(ggplot2))
```

## Set up directories

```{r setup directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
module_dir <- file.path(root_dir, "analyses","molecular-subtyping-NBL")
util_dir <- file.path(module_dir, "util")
plot_dir <- file.path(module_dir, "plots")
subset_dir <- file.path(module_dir, "nbl-subset")
results_dir <- file.path(module_dir, "results")

```

# Plot function

```{r plot chromosome 2 segment means}
# source function to plot chromosome 2  biospecimen segment means
source(file.path(module_dir, "util", "plot_chr2.R"))

```

Load input files and initialize expression threshold base on the threshold 
inferred from MYCN gene expression distribution on plot 
(`plots/TPM_Biospecimen_All_Samples_With_TMP.png`)

## Load files

```{r message=FALSE} 
# biospecimen for subtyping
alteration_df <- 
  readr::read_tsv(file.path(subset_dir, 
                            "alteration_table_without_molecular_subtype.tsv"),
                                    guess_max = 100000)
# cnvkit segment means
cnv_cnvkit_df <- 
  readr::read_tsv(file.path(data_dir, "cnv-cnvkit.seg.gz"), 
                  guess_max = 100000)
# base histologies
hist_df <- 
  readr::read_tsv(file.path(data_dir, "histologies-base.tsv"), 
                  guess_max = 100000)

# tmp expression cutoff 
Suggested_Cutoff <- 140.83 # TARGET-30-PAMEZH-01A-01R	

# MYCN clinical status
mycn_clincial_status <- alteration_df %>% 
  dplyr::select(Kids_First_Biospecimen_ID_DNA, Kids_First_Biospecimen_ID_RNA,
                pathology_free_text_diagnosis) %>% 
  dplyr::rename(molecular_subtype = pathology_free_text_diagnosis)

```

Case 1:
If `pathology_free_text_diagnosis` is `MYCN amp` and `MYCN_CN_status` is 
`amplification`, assign `molecular_subtype` as `NBL, MYCN amplified`

Case 2:
If `pathology_free_text_diagnosis` is `MYCN non-amp` and `MYCN_CN_status` is 
`amplification`,  assign `molecular_subtype` as `NBL, MYCN amplified`

Case 3:
If `pathology_free_text_diagnosis` is `MYCN non-amp` and `MYCN_CN_status` is 
NOT `amplification`, assign `molecular_subtype` as "NBL, MYCN non-amplified"

## Subtype based on clinical and copy number status

```{r subtype based on clinical and copy number}
# assign subtypes for cases 1, 2, 3 described above
mycn_path_free <- 
  c("MYCN non-amp", "ganglioneuroblastoma, stage ii favorable histology non n-myc amplified")
mycn_status <- c("gain","loss","neutral")
alteration_df <- alteration_df %>% 
  dplyr::mutate(molecular_subtype = 
           case_when(
             MYCN_CN_status == "amplification" & 
               pathology_free_text_diagnosis == "MYCN amp" ~ "NBL, MYCN amplified",
             MYCN_CN_status =="amplification" & 
               pathology_free_text_diagnosis %in% mycn_path_free ~ "NBL, MYCN amplified",
             MYCN_CN_status %in% mycn_status &  
               pathology_free_text_diagnosis %in% mycn_path_free ~ "NBL, MYCN non-amplified",
             TRUE ~ NA_character_)
         )

```

Case 4: 
If `pathology_free_text_diagnosis` is `MYCN amp` and `MYCN_CN_status` is 
NOT `amplification`, plot `segment mean vs location` to determine subtype. 

If MYCN is called not amplified (i,e., NA, gain, loss, neutral), but the 
clinical data suggests MYCN amplification, plot the CNV data to visualize 
whether we see focal amplification despite CNV not being called

If `pathology_free_text_diagnosis` is `MYCN amp` MYCN_CN_status is 
NOT `amplification`, and the DNA biospecimen has corresponding RNA biospecimen,
check if the RNA biospecimen has a TPM value and assign amplification status 
based on the `Suggested_Cutoff` inferred from the MYCN expression distribution 
bar plot. 

In cases where there are no TPM values, assign the `molecular_subtype` columns the 
the string values, `Pathology-amp,MYCN_CN_status-non-amp,TPM-NA`. This coding
we eventually be change `NA` in the subsequent steps.

## Subtype based on segment means and TPM

```{r subtype based on segment means and tpm} 
clincal_amplified_IDs <- alteration_df %>% 
  dplyr::filter(MYCN_CN_status %in% mycn_status, 
                pathology_free_text_diagnosis == "MYCN amp") %>% 
  dplyr::pull(Kids_First_Biospecimen_ID_DNA)

# Iterate through the biospecimen list in clincal_amplified_IDs which belong 
# to case 4A and plot the segment means the MYCN chromosome 2 cytoband region
for (iter in 1: length(clincal_amplified_IDs)) {
  plt_file <- 
    file.path(plot_dir, paste(clincal_amplified_IDs[iter],"_chr2p",".png",sep=""))
  # For CI issues
  if(length(clincal_amplified_IDs) > 0) {  
    # Check if clincal_amplified_IDs is NA,
    # which will happen for biospecimen ID which are RNA IDs
    if(!is.na(clincal_amplified_IDs[iter])) {
      plt1 <- plot_chr2(cnv_cnvkit_df,clincal_amplified_IDs[iter])
      if(plt1@data@elementMetadata@nrows > 0) { # CI fail issues being addressed in this LINE
        ggsave(filename = plt_file,plot =plt1@ggplot)
      }
    } else {
      # If clincal_amplified_IDs is NA we need to find corresponding RNA ID
      plt1 <- 
        plot_chr2(cnv_cnvkit_df, 
                  alteration_df$Kids_First_Biospecimen_ID_RNA[index_clinical_amplified[iter]])
      if(plt1@data@elementMetadata@nrows > 0) { # CI fail issues being addressed in this LINE
        ggsave(filename = plt_file,plot =plt1@ggplot)
      }
    }
  }  
}

clincal_amplified_df <- alteration_df %>%
  dplyr::filter(Kids_First_Biospecimen_ID_DNA %in% clincal_amplified_IDs) %>% 
  dplyr::mutate(molecular_subtype =
                  case_when(
                    is.na(MYCN_TPM) ~ "Pathology-amp,MYCN_CN_status-non-amp,TPM-NA",
                    MYCN_TPM >= Suggested_Cutoff ~ "NBL, MYCN amplified",
                    MYCN_TPM < Suggested_Cutoff ~ "NBL, MYCN non-amplified",
                    TRUE ~ as.character(molecular_subtype)
                  )
               )

alteration_df <- alteration_df %>% 
  dplyr::filter(!Kids_First_Biospecimen_ID_DNA %in% clincal_amplified_IDs) %>%
  dplyr::bind_rows(clincal_amplified_df)

```

Case 5: 
If there are biospecimen which don't belong to cases 1-4, but have a TPM value,  
assign molecular subtype based on `Suggested_Cutoff`. 

## Subtype based on TPM

```{r subtype based tpm}
case_5_ids <- alteration_df %>% 
  dplyr::filter(is.na(molecular_subtype), 
                      MYCN_TPM >= Suggested_Cutoff | MYCN_TPM < Suggested_Cutoff) %>% 
  dplyr::pull(Kids_First_Biospecimen_ID_DNA)
                  

alteration_df <- alteration_df %>% 
  dplyr::mutate(molecular_subtype =
                  case_when(
                    (is.na(molecular_subtype) & 
                      MYCN_TPM >= Suggested_Cutoff) ~ "NBL, MYCN amplified",
                    (is.na(molecular_subtype) & 
                      MYCN_TPM < Suggested_Cutoff) ~ "NBL, MYCN non-amplified",
                    TRUE ~ as.character(molecular_subtype)
                  )
               )

```

Case 6: 
If there are biospecimen not in cases 1-5, assign the subtype as `NBL, to be classified`

## Biosepcimen not subtyped

```{r subtype to be classified}
alteration_df <- alteration_df %>% 
  dplyr::mutate(molecular_subtype =
                  case_when(
                    is.na(molecular_subtype) ~ "NBL, to be classified",
                    TRUE ~ as.character(molecular_subtype)
                  )
               ) 

```

Create a dataframe containing all the samples whose molecular_subtype was 
determined based on TPM.

## Biosepcimen subtyped by TMP

```{r df subtyped by tmp}
 alteration_tmp_df <- alteration_df %>% 
  dplyr::filter(Kids_First_Biospecimen_ID_DNA %in% 
                  unique(c(clincal_amplified_IDs, case_5_ids))) %>% 
  dplyr::select(Kids_First_Biospecimen_ID_DNA, 
                Kids_First_Biospecimen_ID_RNA, 
                MYCN_CN,
                pathology_free_text_diagnosis, 
                MYCN_CN_status, MYCN_TPM,
                molecular_subtype)

```


## Final molecular subtyping table

```{r final molecular subtyping table}
# create the final molecular subtyping table 
molecular_subtype_table <- alteration_df %>% 
  dplyr::select(Kids_First_Biospecimen_ID_DNA, 
                Kids_First_Biospecimen_ID_RNA, 
                molecular_subtype) %>% 
  dplyr::mutate(molecular_subtype = 
                  case_when(
                    molecular_subtype == 
                      "Pathology-amp,MYCN_CN_status-non-amp,TPM-NA" ~
                      "NBL, to be classified",
                    TRUE ~ as.character(molecular_subtype)
                  )
               ) %>% 
  dplyr::arrange(molecular_subtype)

mycn_subtyped <- molecular_subtype_table %>% 
  dplyr::filter(molecular_subtype != "NBL, to be classified")

mycn_not_subtyped <- molecular_subtype_table %>% 
  dplyr::filter(molecular_subtype == "NBL, to be classified") %>% 
  dplyr::select(-molecular_subtype) %>% 
  dplyr::left_join(mycn_clincial_status, 
                   by = c("Kids_First_Biospecimen_ID_DNA",
                          "Kids_First_Biospecimen_ID_RNA")) %>% 
  dplyr::mutate(molecular_subtype =
                  case_when(
                    molecular_subtype == "MYCN amp" ~ "NBL, MYCN amplified",
                    molecular_subtype == "MYCN non-amp" ~ "NBL, MYCN non-amplified",
                    TRUE ~ "NBL, to be classified"
                  )
               )

molecular_subtype_table <- bind_rows(mycn_subtyped, mycn_not_subtyped)

```

## Write molecular subtyping table to file

```{r write to file}
alteration_df %>% 
  readr::write_tsv(file.path(results_dir, 
                             "alteration_table_with_molecular_subtype.tsv"))
alteration_tmp_df %>% 
  readr::write_tsv(file.path(results_dir, 
                             "molecular_subtypes_based_on_cutoff.tsv"))
molecular_subtype_table %>% 
  readr::write_tsv(file.path(results_dir, 
                             "neuroblastoma_molecular_subtypes.tsv"))
```

## Session Info

```{r sessioninfo}
sessionInfo() 
```
