---
title: "PedcBio CNV file preparation"
output: html_document
---

```{r load library}
library(tidyverse)
library(readr)
```

### Define directories
```{r define directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analylses", "pedcbio-cnv-prepare")
results_dir <- file.path(analysis_dir, "results")

if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

```

### Read in files
```{r read in files}
# wgs + wxs autosomes
consensus_cnvkit_auto <- readr::read_tsv(file.path(data_dir, "consensus_wgs_plus_cnvkit_wxs_autosomes.tsv.gz"))
# wgs + wxs x and y
consensus_cnvkit_xy <- readr::read_tsv(file.path(data_dir, "consensus_wgs_plus_cnvkit_wxs_x_and_y.tsv.gz"))

# wxs cnvkit seg file
cnvkit_seg <- readr::read_tsv(file.path(data_dir, "cnv-cnvkit.seg.gz"))
# wgs consensus seg file
consensus_seg <- readr::read_tsv(file.path(data_dir, "cnv-consensus.seg.gz"))

# histology file 
histology_df <- readr::read_tsv(file.path(data_dir, "histologies.tsv"), guess_max = 100000)

```

### Filter consensus and cnvkit annotated files 
```{r filter seg files}
# get bs ids of WGS samples
wgs_bsid <- histology_df %>% 
  dplyr::filter(sample_type == "Tumor") %>% 
  dplyr::filter(experimental_strategy == "WGS" ) %>%
  pull(Kids_First_Biospecimen_ID) %>% 
  unique()

# filter consensus to contain only WGS samples
consensus_seg <- consensus_seg %>% 
  filter(ID %in% wgs_bsid)

# get bs ids of WXS samples
wxs_bsid <- histology_df %>% 
  dplyr::filter(sample_type == "Tumor") %>% 
  dplyr::filter(experimental_strategy == "WXS" ) %>%
  pull(Kids_First_Biospecimen_ID) %>% 
  unique()

# filter cnvkit to contain only WXS samples
cnvkit_seg <- cnvkit_seg %>% 
  filter(ID %in% wxs_bsid)
```

# Fix WGS samples first
## Get BS Ids that are not in the annotation file and BS ids in annotation file
```{r}
# BS IDs in consensus seg file
consensus_seg_bs_id <- consensus_seg %>% pull(ID) %>% unique()
rm(consensus_seg)

# filter combined cnv annotated files to contain only WGS in consensus seg file 
consensus_auto  <- consensus_cnvkit_auto %>% 
  dplyr::filter(biospecimen_id %in% consensus_seg_bs_id)
consensus_xy <- consensus_cnvkit_xy %>%
  dplyr::filter(biospecimen_id %in% consensus_seg_bs_id)

# get BS IDs in annotated cnv
annotated_consensus_auto_bs_id <- consensus_auto %>% pull(biospecimen_id) %>% unique()
annotated_consensus_xy_bs_id <- consensus_xy %>% pull(biospecimen_id) %>% unique()

# get BS Ids not in annotated file
diff_xy_bsid <- consensus_seg_bs_id[!consensus_seg_bs_id %in% annotated_consensus_xy_bs_id ]
diff_auto_bsid <- consensus_seg_bs_id[!consensus_seg_bs_id %in% annotated_consensus_auto_bs_id ]

```

## Deal with X and Y first (WGS samples)
### add neutral segs for samples that are not in the annotation file 
```{r}
# get unique gene_symbol + ensembl_id + cytoband
unique_x_set <- consensus_xy %>% 
  dplyr::select(ensembl, gene_symbol, cytoband) %>% 
  filter(grepl("X", cytoband)) %>% 
  distinct()

unique_y_set <- consensus_xy %>% 
  dplyr::select(ensembl, gene_symbol, cytoband) %>% 
  filter(grepl("Y", cytoband)) %>% 
  distinct()

# for samples that are not in the annotation - add neutral for all unique segments
for (i in 1:length(diff_xy_bsid)){
  sample_set <- data.frame()
  sample_added <- diff_xy_bsid[i] 
  sample_gender <- histology_df %>% 
    filter(Kids_First_Biospecimen_ID == sample_added) %>% 
    pull(germline_sex_estimate)
  
  # add back segments based on gender 
  if(sample_gender == "Female") {
    sample_x_set <- unique_x_set %>% 
      mutate(biospecimen_id = sample_added) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 2) %>% 
      mutate(ploidy = 2)
    sample_y_set <- unique_y_set %>% 
      mutate(biospecimen_id = sample_added) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 0) %>% 
      mutate(ploidy = 0)
    sample_set <- bind_rows(sample_x_set, sample_y_set)
  }
  
  if(sample_gender == "Male") {
    sample_x_set <- unique_x_set %>% 
      mutate(biospecimen_id = sample_added) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 1) %>% 
      mutate(ploidy = 1)
    sample_y_set <- unique_y_set %>% 
      mutate(biospecimen_id = sample_added) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 1) %>% 
      mutate(ploidy = 1)
    sample_set <- bind_rows(sample_x_set, sample_y_set)
  }
  consensus_xy  <- bind_rows(consensus_xy, sample_set)
}
```

### check to make sure diff samples now have nrow of unique x+y set 
```{r}
nrow(unique_x_set)
nrow(unique_y_set)
sum(nrow(unique_x_set), nrow(unique_y_set))

consensus_xy %>% filter(biospecimen_id %in% diff_xy_bsid) %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```

### add neutral segs to un-annotated segments for samples that are in the annotation file 
```{r}
for (j in 1:length(annotated_consensus_xy_bs_id)){
  sample_set <- data.frame()
  sample_of_interest <- annotated_consensus_xy_bs_id[j]
  
  # find the gender of the sample
  sample_gender <- histology_df %>% 
    filter(Kids_First_Biospecimen_ID == sample_of_interest) %>% 
    pull(germline_sex_estimate)
  
  # find all the segments that have information for that sample
  sample_annotated_ensembl <- consensus_xy %>% 
    dplyr::filter(biospecimen_id == sample_of_interest) %>% 
    dplyr::pull(ensembl) %>% unique()
  
  # find the segments that do not have annotation
  diff_x_seg <- unique_x_set %>% 
    filter(!ensembl %in% sample_annotated_ensembl)
  diff_y_seg <- unique_y_set %>% 
    filter(!ensembl %in% sample_annotated_ensembl)
  
  # add back segments based on gender 
  if(sample_gender == "Female") {
    sample_x_set <- diff_x_seg %>% 
      mutate(biospecimen_id = sample_of_interest) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 2) %>% 
      mutate(ploidy = 2)
    sample_y_set <- diff_y_seg %>% 
      mutate(biospecimen_id = sample_of_interest) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 0) %>% 
      mutate(ploidy = 0)
    sample_set <- bind_rows(sample_x_set, sample_y_set)
  }
  
  if(sample_gender == "Male") {
    sample_x_set <- diff_x_seg %>% 
      mutate(biospecimen_id = sample_of_interest) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 1) %>% 
      mutate(ploidy = 1)
    sample_y_set <- diff_y_seg %>% 
      mutate(biospecimen_id = sample_of_interest) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 1) %>% 
      mutate(ploidy = 1)
    sample_set <- bind_rows(sample_x_set, sample_y_set)
  }
  consensus_xy  <- bind_rows(consensus_xy, sample_set)
}

```

### check to make sure all samples now have ~nrow of unique x+y set 
```{r}
nrow(unique_x_set)
nrow(unique_y_set)
sum(nrow(unique_x_set), nrow(unique_y_set))

consensus_xy %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```

### check to see what those samples are 
```{r}
consensus_xy %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>%
  filter(count == 7) %>%
  knitr::kable()

consensus_xy %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>%
  filter(count == 11) %>%
  knitr::kable()
```
These have germline_sex_estimat as Unknown - cannot add segments back - expected.

## Deal with autosomes next (WGS samples)
### add neutral segs for samples that are not in the annotation file 
```{r}
# get unique gene_symbol + ensembl_id + cytoband
unique_auto_set <- consensus_auto %>% 
  dplyr::select(ensembl, gene_symbol, cytoband) %>% 
  distinct()
# for BS IDs not in annotation, add all segments back
for (m in 1:length(diff_auto_bsid)){
  sample_set <- data.frame()
  sample_added_auto <- diff_auto_bsid[m]
  sample_set <- unique_auto_set %>% 
    mutate(biospecimen_id = sample_added_auto) %>% 
    mutate(status = "neutral") %>% 
    mutate(copy_number = 2) %>% 
    mutate(ploidy = 2)
  consensus_auto <- bind_rows(consensus_auto, sample_set)
}
```

### check to make sure diff samples now have nrow of unique x+y set 
```{r}
nrow(unique_auto_set)

consensus_auto %>% filter(biospecimen_id %in% diff_auto_bsid) %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```

### add neutral segs to un-annotated segments for samples that are in the annotation file 
```{r}
for (n in 1:length(annotated_consensus_auto_bs_id)){
  sample_set <- data.frame()
  sample_of_interest_auto <- annotated_consensus_auto_bs_id[n]
  sample_annotated_ensembl <- consensus_auto %>% 
    dplyr::filter(biospecimen_id == sample_of_interest_auto) %>% 
    dplyr::pull(ensembl) %>% unique()
  
  # find the segments that do not have annotation
  diff_auto_seg <- unique_auto_set %>% 
    dplyr::filter(!ensembl %in% sample_annotated_ensembl)
  
  sample_set <- diff_auto_seg %>% 
    dplyr::mutate(biospecimen_id = sample_of_interest_auto) %>% 
    dplyr::mutate(status = "neutral") %>% 
    dplyr::mutate(copy_number = 2) %>% 
    dplyr::mutate(ploidy = 2)
  consensus_auto <- bind_rows(consensus_auto, sample_set)
}
```

### check to make sure all samples now have ~ nrow of unique x+y set 
```{r}
nrow(unique_auto_set)

consensus_auto %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```


# Fix WXS samples next
## Get BS Ids that are not in the annotation file and BS ids in annotation file
```{r filter cnvkit annotated files}
# BS IDs in consensus seg file
cnvkit_seg_bs_id <- cnvkit_seg %>% pull(ID) %>% unique()
rm(cnvkit_seg)

# filter combined cnv annotated files to contain only WXS in consensus seg file 
cnvkit_auto  <- consensus_cnvkit_auto %>% 
  dplyr::filter(biospecimen_id %in% cnvkit_seg_bs_id)
cnvkit_xy <- consensus_cnvkit_xy %>%
  dplyr::filter(biospecimen_id %in% cnvkit_seg_bs_id)

# get BS IDs in annotated cnv
annotated_cnvkit_auto_bs_id <- cnvkit_auto %>% pull(biospecimen_id) %>% unique()
annotated_cnvkit_xy_bs_id <- cnvkit_xy %>% pull(biospecimen_id) %>% unique()

# get BS Ids not in annotated file
diff_xy_bsid <- cnvkit_seg_bs_id[!cnvkit_seg_bs_id %in% annotated_cnvkit_xy_bs_id ]
diff_auto_bsid <-cnvkit_seg_bs_id[!cnvkit_seg_bs_id %in% annotated_cnvkit_auto_bs_id ]

```

## Deal with X and Y first (WXS samples)
### add neutral segs for samples that are not in the annotation file 
```{r}
 # get unique gene_symbol + ensembl_id + cytoband
unique_x_set <- cnvkit_xy %>% 
  dplyr::select(ensembl, gene_symbol, cytoband) %>% 
  filter(grepl("X", cytoband)) %>% 
  distinct()

unique_y_set <- cnvkit_xy %>% 
  dplyr::select(ensembl, gene_symbol, cytoband) %>% 
  filter(grepl("Y", cytoband)) %>% 
  distinct()
  
if(length(diff_xy_bsid)>0){
  # for samples that are not in the annotation - add neutral for all unique segments
  for (i in 1:length(diff_xy_bsid)){
    sample_set <- data.frame()
    sample_added <- diff_xy_bsid[i] 
    # gender uses reported gender
    sample_gender <- histology_df %>% 
      filter(Kids_First_Biospecimen_ID == sample_added) %>% 
      pull(reported_gender)
    
    # we can only add back if there is reported gender
    if(sample_gender %in% c("Male", "Female")){
      # add back segments based on gender 
      if(sample_gender == "Female") {
        sample_x_set <- unique_x_set %>% 
          mutate(biospecimen_id = sample_added) %>% 
          mutate(status = "neutral") %>% 
          mutate(copy_number = 2) %>% 
          mutate(ploidy = 2)
        sample_y_set <- unique_y_set %>% 
          mutate(biospecimen_id = sample_added) %>% 
          mutate(status = "neutral") %>% 
          mutate(copy_number = 0) %>% 
          mutate(ploidy = 0)
        sample_set <- bind_rows(sample_x_set, sample_y_set)
      }
      
      if(sample_gender == "Male") {
        sample_x_set <- unique_x_set %>% 
          mutate(biospecimen_id = sample_added) %>% 
          mutate(status = "neutral") %>% 
          mutate(copy_number = 1) %>% 
          mutate(ploidy = 1)
        sample_y_set <- unique_y_set %>% 
          mutate(biospecimen_id = sample_added) %>% 
          mutate(status = "neutral") %>% 
          mutate(copy_number = 1) %>% 
          mutate(ploidy = 1)
        sample_set <- bind_rows(sample_x_set, sample_y_set)
      }
    }
    cnvkit_xy  <- bind_rows(cnvkit_xy, sample_set)
  }
}

```

### check to make sure diff samples now have nrow of unique x+y set 
```{r}
nrow(unique_x_set)
nrow(unique_y_set)
sum(nrow(unique_x_set), nrow(unique_y_set))

cnvkit_xy %>% filter(biospecimen_id %in% diff_xy_bsid) %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```

### add neutral segs to un-annotated segments for samples that are in the annotation file 
```{r}
for (j in 1:length(annotated_cnvkit_xy_bs_id)){
  sample_set <- data.frame()
  sample_of_interest <- annotated_cnvkit_xy_bs_id[j]
  
  # find the gender of the sample
  sample_gender <- histology_df %>% 
    filter(Kids_First_Biospecimen_ID == sample_of_interest) %>% 
    pull(reported_gender)
  
  # only add back when reported gender is known
  if(sample_gender %in% c("Male", "Female")){
    # find all the segments that have information for that sample
    sample_annotated_ensembl <- cnvkit_xy %>% 
      dplyr::filter(biospecimen_id == sample_of_interest) %>% 
      dplyr::pull(ensembl) %>% unique()
    
    # find the segments that do not have annotation
    diff_x_seg <- unique_x_set %>% 
      filter(!ensembl %in% sample_annotated_ensembl)
    diff_y_seg <- unique_y_set %>% 
      filter(!ensembl %in% sample_annotated_ensembl)
    
    # add back segments based on gender 
    if(sample_gender == "Female") {
      sample_x_set <- diff_x_seg %>% 
        mutate(biospecimen_id = sample_of_interest) %>% 
        mutate(status = "neutral") %>% 
        mutate(copy_number = 2) %>% 
        mutate(ploidy = 2)
      sample_y_set <- diff_y_seg %>% 
        mutate(biospecimen_id = sample_of_interest) %>% 
        mutate(status = "neutral") %>% 
        mutate(copy_number = 0) %>% 
        mutate(ploidy = 0)
      sample_set <- bind_rows(sample_x_set, sample_y_set)
    }
    
    if(sample_gender == "Male") {
      sample_x_set <- diff_x_seg %>% 
        mutate(biospecimen_id = sample_of_interest) %>% 
        mutate(status = "neutral") %>% 
        mutate(copy_number = 1) %>% 
        mutate(ploidy = 1)
      sample_y_set <- diff_y_seg %>% 
        mutate(biospecimen_id = sample_of_interest) %>% 
        mutate(status = "neutral") %>% 
        mutate(copy_number = 1) %>% 
        mutate(ploidy = 1)
      sample_set <- bind_rows(sample_x_set, sample_y_set)
    } 
    cnvkit_xy  <- bind_rows(cnvkit_xy, sample_set)
  }
}

```


### check to make sure all samples now have ~nrow of unique x+y set 
```{r}
nrow(unique_x_set)
nrow(unique_y_set)
sum(nrow(unique_x_set), nrow(unique_y_set))

cnvkit_xy %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```

## Combine consensus and cnvkit for XY together
```{r}
# combine xy 
consensus_cnvkit_xy <- bind_rows(consensus_xy, cnvkit_xy)

# write out results 
consensus_cnvkit_xy %>% 
  readr::write_tsv("results/consensus_wgs_plus_cnvkit_wxs_x_and_y.tsv.gz")

rm(consensus_xy)
rm(cnvkit_xy)

```

## Deal with autosomes next (WXS samples)
### add neutral segs for samples that are not in the annotation file 
```{r}
# get unique gene_symbol + ensembl_id + cytoband
unique_auto_set <- cnvkit_auto %>% 
  dplyr::select(ensembl, gene_symbol, cytoband) %>% 
  distinct()
# for BS IDs not in annotation, add all segments back
if(length(diff_auto_bsid)>0){
  for (m in 1:length(diff_auto_bsid)){
    sample_set <- data.frame()
    sample_added_auto <- diff_auto_bsid[m]
    sample_set <- unique_auto_set %>% 
      mutate(biospecimen_id = sample_added_auto) %>% 
      mutate(status = "neutral") %>% 
      mutate(copy_number = 2) %>% 
      mutate(ploidy = 2)
    cnvkit_auto <- bind_rows(cnvkit_auto, sample_set)
  }
}

```

### check to make sure diff samples now have nrow of unique x+y set 
```{r}
nrow(unique_auto_set)

cnvkit_auto %>% filter(biospecimen_id %in% diff_auto_bsid) %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```

### add neutral segs to un-annotated segments for samples that are in the annotation file 
```{r}
for (n in 1:length(annotated_cnvkit_auto_bs_id)){
  sample_set <- data.frame()
  sample_of_interest_auto <- annotated_cnvkit_auto_bs_id[n]
  sample_annotated_ensembl <- cnvkit_auto %>% 
    dplyr::filter(biospecimen_id == sample_of_interest_auto) %>% 
    dplyr::pull(ensembl) %>% unique()
  
  # find the segments that do not have annotation
  diff_auto_seg <- unique_auto_set %>% 
    dplyr::filter(!ensembl %in% sample_annotated_ensembl)
  
  sample_set <- diff_auto_seg %>% 
    dplyr::mutate(biospecimen_id = sample_of_interest_auto) %>% 
    dplyr::mutate(status = "neutral") %>% 
    dplyr::mutate(copy_number = 2) %>% 
    dplyr::mutate(ploidy = 2)
  cnvkit_auto <- bind_rows(cnvkit_auto, sample_set)
}
```

### check to make sure all samples now have ~ nrow of unique x+y set 
```{r}
nrow(unique_auto_set)

cnvkit_auto %>% 
  group_by(biospecimen_id) %>%
  mutate(count = n()) %>% 
  pull(count) %>% 
  unique()

```


## Combine autosomes together
```{r}
# combine autosomes
consensus_cnvkit_auto <- bind_rows(consensus_auto, cnvkit_auto)

# write out results
consensus_cnvkit_auto %>% 
  readr::write_tsv("results/consensus_wgs_plus_cnvkit_wxs_autosomes.tsv.gz")

rm(consensus_auto)
rm(cnvkit_auto)
```

## Combine cnv and autosomes and add germline sex estimate
```{r}
# first remove germline sex estimate 
consensus_cnvkit_xy <- consensus_cnvkit_xy %>%
  select(-germline_sex_estimate)

# get the germline sex estimate for all samples
germline_sex_estimate_df <- histology_df %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% consensus_seg_bs_id) %>% 
  dplyr::rename(biospecimen_id = Kids_First_Biospecimen_ID) %>% 
  dplyr::mutate(gender = case_when(
    !is.na(germline_sex_estimate) ~ germline_sex_estimate,
    is.na(germline_sex_estimate) ~ reported_gender
  )) %>% 
  dplyr::select(biospecimen_id, gender)%>% 
  dplyr::rename(germline_sex_estimate = gender)

# annotate germline sex estimate of those samples and write out the results
bind_rows(consensus_cnvkit_xy, consensus_cnvkit_auto) %>% 
  left_join(germline_sex_estimate_df) %>%
  distinct() %>% 
  write_tsv("results/consensus_wgs_plus_cnvkit_wxs.tsv.gz")

```

