---
title: "Identify samples with _TP53_ and _NF1_ mutations in the stranded RNA-seq dataset"
output: html_notebook
author: Jaclyn Taroni for ALSF CCDL
date: 2020
---

To run and fully test the `tp53_nf1_score` module in continuous integration, we must ensure that there are positive examples of _TP53_ and _NF1_ mutations.
Furthermore, we need to include the same samples in the RNA-seq data.
Because we skip running the `tp53_nf1_score` classifier and evaluation steps on the poly-A dataset in CI, we can focus on the stranded data.

```{r}
library(tidyverse)
```

```{r}
set.seed(2022)
```

The table prepped in `tp53_nf1_score` contains all of the information we need regarding positive examples.
The histologies file contains the information we need for identifying samples that are in the stranded RNA-seq dataset.

```{r}
mutations_table <- read_tsv(file.path("..", "tp53_nf1_score", "results", 
                                      "TP53_NF1_snv_alteration.tsv"))
histologies_df <- read_tsv(file.path("..", "..", "data", 
                                     "histologies.tsv"),
                           col_types = cols(
                             molecular_subtype = col_character()
                           ))
```

Get biospecimen identifiers for samples with mutations of interest.

```{r}
tp53_examples <- mutations_table %>%
  filter(Hugo_Symbol == "TP53") %>%
  pull(Tumor_Sample_Barcode)
nf1_examples <- mutations_table %>%
  filter(Hugo_Symbol == "NF1") %>%
  pull(Tumor_Sample_Barcode)
```

### _TP53_ positive examples

```{r}
# samples that also have RNA-seq data
tp53_samples <- histologies_df %>%
  filter(Kids_First_Biospecimen_ID %in% tp53_examples) %>%
  group_by(group_id = paste(sample_id, Kids_First_Participant_ID, sep ="-")) %>%
  ungroup() %>% 
  inner_join(histologies_df %>%
               filter(RNA_library == "stranded" | 
                        RNA_library == "poly-A stranded" |
                        RNA_library == "poly-A",
                      sample_type == "Tumor", 
                      experimental_strategy != "Methylation") %>%
               group_by(group_id = paste(sample_id, 
                                         Kids_First_Participant_ID, sep ="-")) %>%
               ungroup() %>% 
               select(group_id)) %>%
  distinct()
```

#### PBTA _TP53_ positive examples

```{r}
# randomly selected PBTA positive examples
histologies_df %>%
  filter(cohort == "PBTA",
         sample_type == "Tumor",
         experimental_strategy != "Methylation") %>% 
  group_by(group_id = paste(sample_id, 
                            Kids_First_Participant_ID, sep ="-")) %>% 
  ungroup() %>% 
  filter(group_id %in% sample(tp53_samples %>% 
                                 filter(cohort == "PBTA") %>% 
                                 pull(group_id) %>% unique(),
                                 size = 3)
         ) %>% 
  arrange(group_id)
```

#### GMKF _TP53_ positive examples

```{r}
# randomly selected GMKF positive examples
histologies_df %>%
  filter(cohort == "GMKF",
         sample_type == "Tumor",
         experimental_strategy != "Methylation") %>% 
  group_by(group_id = paste(sample_id, 
                            Kids_First_Participant_ID, sep ="-")) %>% 
  ungroup() %>% 
  filter(group_id %in% sample(tp53_samples %>% 
                                 filter(cohort == "GMKF") %>% 
                                 pull(group_id) %>% unique(),
                                 size = 1)
         ) %>% 
  arrange(group_id)
```

#### TARGET _TP53_ positive examples

```{r}
# randomly selected TARGET positive examples
histologies_df %>%
  filter(cohort == "TARGET",
         sample_type == "Tumor",
         experimental_strategy != "Methylation") %>% 
  group_by(group_id = paste(sample_id, 
                            Kids_First_Participant_ID, sep ="-")) %>% 
  ungroup() %>% 
  filter(group_id %in% sample(tp53_samples %>% 
                                 filter(cohort == "TARGET") %>% 
                                 pull(group_id) %>% unique(),
                                 size = 3)
         ) %>% 
  arrange(group_id)
```

### _NF1_ positive examples

```{r}
# randomly select samples that also have stranded RNA-seq data
# samples that also have RNA-seq data
nf1_samples <- histologies_df %>%
  filter(Kids_First_Biospecimen_ID %in% nf1_examples) %>%
  group_by(group_id = paste(sample_id, Kids_First_Participant_ID, sep ="-")) %>%
  ungroup() %>% 
  inner_join(histologies_df %>%
               filter(RNA_library == "stranded" | 
                        RNA_library == "poly-A stranded" |
                        RNA_library == "poly-A",
                      sample_type == "Tumor", 
                      experimental_strategy != "Methylation") %>%
               group_by(group_id = paste(sample_id, 
                                         Kids_First_Participant_ID, sep ="-")) %>% 
               ungroup() %>% 
               select(group_id)) %>%
  distinct()
```

#### PBTA _NF1_ positive examples

```{r}
# randomly selected PBTA positive examples
histologies_df %>%
  filter(cohort == "PBTA",
         sample_type == "Tumor",
         experimental_strategy != "Methylation") %>% 
  group_by(group_id = paste(sample_id, 
                            Kids_First_Participant_ID, sep ="-")) %>% 
  ungroup() %>%
  filter(group_id %in% sample(nf1_samples %>% 
                                 filter(cohort == "PBTA") %>% 
                                 pull(group_id) %>% unique(),
                                 size = 3)
         ) %>% 
  arrange(group_id)
```

#### GMKF _NF1_ positive examples

```{r}
# randomly selected GMKF positive examples
histologies_df %>%
  filter(cohort == "GMKF",
         sample_type == "Tumor",
         experimental_strategy != "Methylation") %>% 
  group_by(group_id = paste(sample_id, 
                            Kids_First_Participant_ID, sep ="-")) %>% 
  ungroup() %>%
  filter(group_id %in% (nf1_samples %>% 
                                 filter(cohort == "GMKF") %>% 
                                 pull(group_id) %>% unique())
         ) %>% 
  arrange(group_id)
```

#### TARGET _NF1_ positive examples

```{r}
# randomly selected TARGET positive examples
histologies_df %>%
  filter(cohort == "TARGET",
         sample_type == "Tumor",
         experimental_strategy != "Methylation") %>% 
  group_by(group_id = paste(sample_id, 
                            Kids_First_Participant_ID, sep ="-")) %>% 
  ungroup() %>%
  filter(group_id %in% sample(nf1_samples %>% 
                                 filter(cohort == "TARGET") %>% 
                                 pull(group_id) %>% unique(),
                                 size = 3)
         ) %>% 
  arrange(group_id)
```

### Session Info

```{r}
sessionInfo()
```

