---
title: "GSVA Score Modeling"
author: "Stephanie J. Spielman for ALSF CCDL"
date: '2020'
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    df_print: paged
    toc: yes
    toc_float: yes
params:
  plot_ci: yes
  is_ci: FALSE
---

### Purpose

The purpose of this analysis is to assess significant differences in GSVA scores for each hallmark pathways. Using ANOVA and subsequent Tukey tests, we ask:

+ For each pathway, are GSVA scores significantly different across `cancer_group`? If so, which histologies are significantly different?

+ For each pathway, are GSVA scores significantly different across `broad_histology`? If so, which types are significantly different?

We perform this using both GSVA scores calculated from RNA-seq libraries. Code is also flexible enough to test a different variable besides `cancer_group`, etc.

### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/gene-set-enrichment-analysis/02-model-gsea.Rmd', clean = TRUE)" 
```
_This assumes you are in the top directory of the repository._

### Setup

Load libraries and define certain constants:

```{r, lib-load, warning=FALSE, message=FALSE}
library(tidyverse)
library(broom)
`%>%` <- dplyr::`%>%`

# This script contains functions used to modeling GSVA scores
source(file.path("util", "hallmark_models.R"))

# Significance testing universal threshold
SIGNIFICANCE_THRESHOLD <- 0.01

# Assigning params$is_ci to running_in_ci avoids a locked binding error
running_in_ci <- params$is_ci

# Are we testing? In case of a non 0/1 number, we recast as logical, and then ensure logical.
if (running_in_ci %in% c(0,1)) running_in_ci <- as.logical(running_in_ci)
if (!(is.logical(running_in_ci)))
{
  stop("\n\nERROR: The parameter `is_ci` should be FALSE/TRUE (or 0/1).")
}

```


Next, define directories and load data files:
```{r, data-load, message=FALSE, warning=FALSE}
### Define directories
# Establish base dir
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir    <- file.path(root_dir, "data") 
results_dir <- "results"

######### Define input files
## Metadata file (histologies/clinical data)
metadata_file <- file.path(data_dir, "histologies.tsv")

## GSEA scores
scores_file <- file.path(results_dir, "gsva_scores.tsv")

######## Load input files
metadata    <- readr::read_tsv(metadata_file, guess_max = 100000) %>% 
  filter(experimental_strategy == "RNA-Seq") %>% 
  filter(!is.na(RNA_library))

scores_file <- readr::read_tsv(scores_file) 

######## Find out unique RNA library types
rna_library_list <- scores_file %>% pull(data_type) %>% unique()

######## Define output files for each unique rna library
cancer_group_anova_outpaths <- lapply(rna_library_list, function(x){
  x<-gsub(" ", "_", x)
  x<-stringr::str_to_lower(gsub("-", "", x))
  file.path(results_dir, paste0("gsva_anova_", x, "_cancer_group.tsv"))
})

cancer_group_tukey_outpaths <- lapply(rna_library_list, function(x){
  x<-gsub(" ", "_", x)
  x<-stringr::str_to_lower(gsub("-", "", x))
  file.path(results_dir, paste0("gsva_tukey_", x, "_cancer_group.tsv"))
})

broad_histology_anova_outpaths <- lapply(rna_library_list, function(x){
  x<-gsub(" ", "_", x)
  x<-stringr::str_to_lower(gsub("-", "", x))
  file.path(results_dir, paste0("gsva_anova_", x, "_broad_histology.tsv"))
})

broad_histology_tukey_outpaths <- lapply(rna_library_list, function(x){
  x<-gsub(" ", "_", x)
  x<-stringr::str_to_lower(gsub("-", "", x))
  file.path(results_dir, paste0("gsva_tukey_", x, "_broad_histology.tsv"))
})

```


### ANOVA and Tukey analysis of GSVA scores

Here we perform a series of ANOVAs, for polyA and stranded libraries separately, to determine whether mean GSVA scores for a given grouping are significantly different across hallmarks (pathways). The given groupings examined here are `cancer_group` and `broad_histology`. 
In other words, we perform an ANOVA (and associated posthoc Tukey test) for each hallmark as, `hallmark ~ grouping`. Users can specify the grouping variable.

First, prepare the data for modeling:

```{r aov-prep}

### Merge histology metadata with each set of gsea scores

metadata_with_gsva <- metadata %>%
  inner_join(scores_file, by = "Kids_First_Biospecimen_ID") 
```


Now, model:

```{r, aov-perform}
for(i in 1:length(rna_library_list)){
  
  rna_library = rna_library_list[i]
  # find out the number of cancer group with this RNA library
  cancer_group_n <- metadata_with_gsva %>%
    filter(data_type == rna_library) %>%
    pull(cancer_group) %>% 
    unique() %>% length()
  
  # find out the number of harmonized diagnosis with this RNA library
  broad_histology_n <- metadata_with_gsva %>%
    filter(data_type == rna_library) %>%
    pull(broad_histology) %>% 
    unique() %>% length()
  
  # anova can only be run on factors with >=2 levels, so to avoid error, we give a if statement 
  if(cancer_group_n>=2){
    cancer_group_model_results <- gsva_anova_tukey(metadata_with_gsva, cancer_group, rna_library, SIGNIFICANCE_THRESHOLD)
    
    # write out results
    readr::write_tsv(cancer_group_model_results[["anova"]], cancer_group_anova_outpaths[[i]])
    readr::write_tsv(cancer_group_model_results[["tukey"]],  cancer_group_tukey_outpaths[[i]])
    
    # print results for viewing 
    print(rna_library)
    print(head(cancer_group_model_results))
  }
  
  # anova can only be run on factors with >=2 levels, so to avoid error, we give a if statement
  if(broad_histology_n>=2){
  broad_histology_model_results <- gsva_anova_tukey(metadata_with_gsva, broad_histology, rna_library, SIGNIFICANCE_THRESHOLD)
  
  # write out results
  readr::write_tsv(broad_histology_model_results[["anova"]], broad_histology_anova_outpaths[[i]])
  readr::write_tsv(broad_histology_model_results[["tukey"]], broad_histology_tukey_outpaths[[i]])
  
  # print results for viewing
  print(rna_library)
  print(head(broad_histology_model_results))
  }
}

# CI part is commented out since we want to run in CAVATICA later
### Don't run polyA samples in CI due to data limitations, won't be enough levels for ANOVA
# if (!(running_in_ci)){
#   cancer_group_polya_model_results <- gsva_anova_tukey(metadata_with_gsva, cancer_group, 'polya', SIGNIFICANCE_THRESHOLD)
#   broad_histology_polya_model_results <- gsva_anova_tukey(metadata_with_gsva, broad_histology, 'polya', SIGNIFICANCE_THRESHOLD)
#   head(cancer_group_polya_model_results)
#   head(broad_histology_polya_model_results)
# 
#   ### Save polya library results
#   write_tsv(cancer_group_polya_model_results[["anova"]],     file_anova_polya_cancer_group)
#   write_tsv(broad_histology_polya_model_results[["anova"]],    file_anova_polya_broad_histology)
#   write_tsv(cancer_group_polya_model_results[["tukey"]],     file_tukey_polya_cancer_group)
#   write_tsv(broad_histology_polya_model_results[["tukey"]],    file_tukey_polya_broad_histology)
#   }

```

**How many `cancer_group` have significant ANOVAs across hallmark pathways, for each library?**
```{r}

for(i in 1:length(rna_library_list)){
  
  rna_library = rna_library_list[i]
   cancer_group_n <- metadata_with_gsva %>%
    filter(data_type == rna_library) %>%
    pull(cancer_group) %>% 
    unique() %>% length()
  
  if(cancer_group_n>=2 ){
    cancer_group_model_results <- gsva_anova_tukey(metadata_with_gsva, cancer_group, rna_library, SIGNIFICANCE_THRESHOLD)
    num_sig <- cancer_group_model_results[["anova"]] %>% 
      count(significant_anova) 
    
    # print out information
   num_sig
  }
}

# CI part is commented out since we want to run in CAVATICA later
# if (!(running_in_ci))
# {
#   cancer_group_polya_model_results[["anova"]] %>% count(significant_anova)
# }
```

> All are significantly different for stranded, none for polya. Likely due to data/power limitations.

**How many `broad_histology` have significant ANOVAs across hallmark pathways, for each library?**
```{r}
for(i in 1:length(rna_library_list)){
  # model for each library type
  rna_library = rna_library_list[i]
  
  broad_histology_n <- metadata_with_gsva %>%
    filter(data_type == rna_library) %>%
    pull(broad_histology) %>% 
    unique() %>% length()
  
  if(broad_histology_n>=2){
    broad_histology_model_results <- gsva_anova_tukey(metadata_with_gsva, broad_histology, rna_library, SIGNIFICANCE_THRESHOLD)
    # count the number of significant 
    num_sig <- broad_histology_model_results[["anova"]] %>% 
      count(significant_anova) 
  # print out information
  num_sig
  }
}

# CI part is commented out since we want to run in CAVATICA later
# if (!(running_in_ci))
# {
#   broad_histology_polya_model_results[["anova"]] %>% count(significant_anova)
# }
```



