---
title: "PedCBio Sample Name Column Addition"
author: Run Jin
output: html_notebook
---

```{r load library}
library(tidyverse)
library(readr)
```

### Define directories
```{r define directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "pedcbio-sample-name")
input_dir <- file.path(analysis_dir, "input")

results_dir <- file.path(analysis_dir, "results")
if(!dir.exists(results_dir)){
  dir.create(results_dir)
}
```

### Read in files
```{r read in files}
# histology file 
histology_df <- readr::read_tsv(file.path(data_dir, "histologies.tsv"), guess_max = 100000)

# read in all the pedcbio files
cbio_names_list <- list.files(input_dir)

files_list <- lapply(cbio_names_list, function(x){
  x <- readr::read_csv(file.path(input_dir, x))
})
```

### Update the ID for those samples if available 
```{r}
formatted_specimens_list <- lapply(files_list, function(x){

  # format them and select relevant columns
  formatted <- x %>% 
    dplyr::mutate(specimen_id = gsub("\\{", "", specimen_id)) %>% 
    dplyr::mutate(specimen_id = gsub("\\}", "", specimen_id)) %>% 
    dplyr::mutate(Kids_First_Biospecimen_ID = strsplit(specimen_id, ",")) %>% 
    unnest(Kids_First_Biospecimen_ID) %>%
    dplyr::select(Kids_First_Biospecimen_ID, formatted_sample_id)
  
  # add those to histology file
  histology_formatted_added <- histology_df %>%
    dplyr::filter(Kids_First_Biospecimen_ID %in% formatted$Kids_First_Biospecimen_ID) %>%
    left_join(formatted)

  return(histology_formatted_added)
})

# get the combined data
combined_histology_formatted_added <- do.call("rbind", formatted_specimens_list)
```

### Add IDs to those without sample id already
```{r}
# find the samples that do not have an ID yet 
histology_df_no_format_id <- histology_df %>%
  dplyr::filter(!Kids_First_Biospecimen_ID %in% combined_histology_formatted_added$Kids_First_Biospecimen_ID) 

```

#### Handle each cohort at a time - start with PBTA
```{r}
# get all sample IDs in the PBTA cohort
sample_ids_pbta <- histology_df_no_format_id %>% 
  dplyr::filter(cohort == "PBTA") %>% 
  pull(sample_id) %>% 
  unique()

# find the samples that need additional `tie-breaker`
specimens_id_need_tiebreak <- c()

for (i in 1:length(sample_ids_pbta)){
  # deal with one sample at a time
  sample_id_of_interest <- sample_ids_pbta[i]
  
  # find the number of compositions
  each_specimen_need_tiebreak <- histology_df_no_format_id %>% 
    dplyr::filter(sample_type == "Tumor") %>% 
    dplyr::filter(sample_id == sample_id_of_interest) %>% 
    group_by(experimental_strategy) %>% 
    dplyr::mutate(n_sample_type = n()) %>%
    dplyr::filter(n_sample_type >1) %>%
    pull(Kids_First_Biospecimen_ID)
  
  # store the samples with multiple compositions
  specimens_id_need_tiebreak <- c(specimens_id_need_tiebreak, each_specimen_need_tiebreak)
}

# see the samples that need tiebreak
pbta_add_tiebreak <- histology_df_no_format_id %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% specimens_id_need_tiebreak,
                sample_type != "Normal") 
```

For DNA, we can use aliquot ID to separate them
For RNA, different specimens are generated from the same aliqout but used different seuqencing method, hence RNA_library type can be used to separate them 
```{r}
pbta_add_tiebreak <- pbta_add_tiebreak %>% 
  dplyr::mutate(formatted_sample_id = case_when(
    experimental_strategy == "RNA-Seq" ~ paste0(sample_id, "-", RNA_library),
    experimental_strategy != "RNA-Seq" ~ paste0(sample_id, "-", aliquot_id)
  ))
```

### Generate pedcbio ID for GTEx dataset 
The format of the sample ID is GTEX-[donor ID]-[tissue site ID]-SM-[aliquot ID].
The donor ID (e.g. GTEX-14753) should be used to link between the various RNA-seq and genotype samples that come from the same donor.

```{r}
gtex_add_tiebreak <- histology_df_no_format_id %>% 
  dplyr::filter(cohort == "GTEx") %>%
  dplyr::mutate(formatted_sample_id = paste0(Kids_First_Participant_ID, "-", aliquot_id))
  
```

### Generate pedcbio ID for TARGET dataset 
Similar to PBTA cohort, only need to add a tiebreak when multiple samples with the same experimental strategy
The different is - we need to use `Kids_First_Participant_ID`
```{r}
# get all participant IDs
participant_ids_target <- histology_df_no_format_id %>% 
  dplyr::filter(cohort == "TARGET") %>% 
  pull(Kids_First_Participant_ID) %>% 
  unique()

# find the samples that need additional `tie-breaker`
specimens_id_need_tiebreak <- c()

for (i in 1:length(participant_ids_target)){
  # deal with one sample at a time
  participant_of_interest <- participant_ids_target[i]
  
  # find the number of compositions
  each_specimen_need_tiebreak <- histology_df_no_format_id %>% 
    dplyr::filter(sample_type == "Tumor") %>% 
    dplyr::filter(Kids_First_Participant_ID == participant_of_interest) %>% 
    group_by(experimental_strategy) %>% 
    dplyr::mutate(n_sample_type = n()) %>%
    dplyr::filter(n_sample_type >1) %>%
    pull(Kids_First_Biospecimen_ID)
  
  # store the samples with multiple compositions
  specimens_id_need_tiebreak <- c(specimens_id_need_tiebreak, each_specimen_need_tiebreak)
}

# see the samples that need tiebreak
target_add_tiebreak <- histology_df_no_format_id %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% specimens_id_need_tiebreak,
                sample_type != "Normal") 

```

For TARGET cohort, looks like using the last 7 digits from the specimen ID + Participant ID should be enough 
```{r}
target_add_tiebreak <- target_add_tiebreak %>%
  dplyr::mutate(formatted_sample_id = gsub("^.{0,10}", "", Kids_First_Biospecimen_ID))
```

### Generate pedcbio ID for TCGA
TCGA-{TSS}-{Participant}-{Sample Vial}-{Portion Analyte}-{Plate}-{Center}
Again using the same method as PBTA and TARGET

```{r}
# get all participant IDs
participant_ids_tcga <- histology_df_no_format_id %>% 
  dplyr::filter(cohort == "TCGA") %>% 
  pull(Kids_First_Participant_ID) %>% 
  unique()

# find the samples that need additional `tie-breaker`
specimens_id_need_tiebreak <- c()

for (i in 1:length(participant_ids_tcga)){
  # deal with one sample at a time
  participant_of_interest <- participant_ids_tcga[i]
  
  # find the number of compositions
  each_specimen_need_tiebreak <- histology_df_no_format_id %>% 
    dplyr::filter(sample_type == "Tumor") %>% 
    dplyr::filter(Kids_First_Participant_ID == participant_of_interest) %>% 
    group_by(experimental_strategy) %>% 
    dplyr::mutate(n_sample_type = n()) %>%
    dplyr::filter(n_sample_type >1) %>%
    pull(Kids_First_Biospecimen_ID)
  
  # store the samples with multiple compositions
  specimens_id_need_tiebreak <- c(specimens_id_need_tiebreak, each_specimen_need_tiebreak)
}

# see the samples that need tiebreak
tcga_add_tiebreak <- histology_df_no_format_id %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% specimens_id_need_tiebreak,
                sample_type != "Normal") 
```

From what we can see, using Kids_First_Biospecimen_ID plus sample ID and aliquot ID is enough 
```{r}
tcga_add_tiebreak <- tcga_add_tiebreak %>%
  dplyr::mutate(formatted_sample_id = paste0(Kids_First_Participant_ID, "-", sample_id, "-", aliquot_id))
```

### Combine all the samples with tiebreak
```{r}
# combine all tie break needed samples
all_tiebreaks <- bind_rows(combined_histology_formatted_added,
                            pbta_add_tiebreak,
                            gtex_add_tiebreak,
                            target_add_tiebreak,
                            tcga_add_tiebreak)

# for samples no need for tie break - use `sample_id` for PBTA and participant id for the rest
no_need_for_tiebreaks <- histology_df %>%
  dplyr::filter(!Kids_First_Biospecimen_ID %in% all_tiebreaks$Kids_First_Biospecimen_ID) %>%
  dplyr::mutate(formatted_sample_id = case_when(
    cohort == "PBTA" ~ sample_id,
    TRUE ~ Kids_First_Participant_ID
  ))

# combine them together and write out 
bind_rows(all_tiebreaks, no_need_for_tiebreaks) %>% 
  readr::write_tsv(file.path(results_dir, "histologies-formatted-id-added.tsv"))
```

