---
title: "Gene Expression TPM Table Summary and QC Checks"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 4
author: Aditya Lahiri and Eric Wafula
date: 2022-11-07
params:
  current_table:
    label: "current expression tpm table"
    value: current/long_n_tpm_mean_sd_quantile_gene_wise_zscore.tsv.gz
    input: file
  previous_table:
    label: "previous expression tpm table"
    value: previous/long_n_tpm_mean_sd_quantile_gene_wise_zscore.tsv.gz
    input: file
---



### Purpose
Performs summary and QC checks comparing the current and the previous OpenPedCan 
tpm tables.

### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('02-tpm-tables-checks.Rmd', clean = TRUE)"
```
_This assumes you are in the analysis module directory of the repository,_
_OpenPedCan-analysis/analyses/mtp-tables-qc-checks._



### Load packages
```{r load packages}
# R packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(openxlsx))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```


### Set output directory
```{r set output directory}
# directories for input and output files
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analyses_dir <- file.path(root_dir, "analyses", "mtp-tables-qc-checks")
results_dir <- file.path(analyses_dir, "results")

# Create results folder if it doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Functions
```{r functions}
# get number of samples in cohorts
get_num_samples <- function(tpm_df) {
  samples <- tpm_df %>% 
    dplyr::filter(cohort != "All Cohorts") %>%
    dplyr::select(cohort, Disease,
                  n_samples) %>%
    dplyr::distinct() %>% 
    dplyr::select(-Disease) %>%
    dplyr::group_by(cohort) %>% 
    dplyr::summarise(n_samples = 
                       sum(as.integer(n_samples, na.rm = TRUE)))
    return(samples)
}

# changes in common columns among MPT mutation frequencies  between 
# current and previous tables
changes_in_columns <- function(current_table, previous_table, column_name) {
  # values specific to current table
  specific_to_current <- current_table %>% 
  dplyr::select(cohort, column_name) %>% 
  dplyr::distinct() %>% 
    setdiff(previous_table %>% dplyr::select(cohort, column_name) %>%
              dplyr::distinct()) %>% 
    dplyr::rename(current_dataset = cohort)
  # values specific to previous table
  specific_to_previous <- previous_table %>% 
    dplyr::select(cohort, column_name) %>% 
    dplyr::distinct() %>% 
    setdiff(current_table %>% dplyr::select(cohort, column_name) %>%
              dplyr::distinct()) %>% 
    dplyr::rename(previous_dataset = cohort)
  # combine differences 
  changes_in_columns <- specific_to_current %>%
    dplyr::full_join(specific_to_previous, by = column_name) %>% 
    dplyr::select(column_name, current_dataset, previous_dataset)
  return(changes_in_columns)
}
```


### Read in current and previous tpm files
```{r load files}
current_table <- data.table::fread(params$current_table, 
                                   data.table = FALSE, 
                                   showProgress = FALSE) %>% 
  purrr::discard(~all(is.na(.))) 

previous_table <- data.table::fread(params$previous_table, 
                                    data.table = FALSE, 
                                    showProgress = FALSE) %>% 
  purrr::discard(~all(is.na(.))) 
```


### First 50 records from a static cancer group (GMKF Neuroblastoma) 
Records from a static cancer_group should not change between the current and previous mutation frequencies tables. There should be no changes in computed mutation frequencies. But other fields in the table will likely change if annotations are updated. 

#### Current table
```{r current table}
# ordered top 50 GMKF Neuroblastoma records from current tpm table
gmkf_nbl_current <- current_table %>% 
  dplyr::filter(cohort == "GMKF", Disease == "Neuroblastoma") %>% 
  dplyr::arrange() %>% 
  head(n = 50)

# write to file
gmkf_nbl_current %>% 
  readr::write_tsv(file.path(results_dir, "current_group_static_cancer.tsv")) 

# display      
gmkf_nbl_current   
```

#### Previous table

```{r previous table}
# ordered top 50 GMKF Neuroblastoma records from previous tpm table
gmkf_nbl_previous <- previous_table %>% 
  dplyr::filter(cohort == "GMKF", Disease == "Neuroblastoma") %>% 
  dplyr::arrange() %>% 
  head(n = 50)

# write to file
gmkf_nbl_previous %>% 
  readr::write_tsv(file.path(results_dir, "previous_group_static_cancer.tsv"))

# display   
gmkf_nbl_previous  
```



### Number of samples in each cohort 
```{r number of samples}
# get number of samples in each cohort
num_samples <- get_num_samples(current_table) %>% 
  dplyr::rename(samples_in_current = n_samples) %>% 
  dplyr::full_join(get_num_samples(previous_table) %>% 
                    dplyr::rename(samples_in_previous = n_samples), by = "cohort") %>% 
  dplyr::select(cohort, samples_in_current, samples_in_previous)

# write to file
num_samples %>% 
  readr::write_tsv(file.path(results_dir, "number_of_samples.tsv"), na = "NA")

# display   
num_samples
```



### Cancer groups in the "All Cohorts" category
```{r all cohorts cancer groups}
# get All Cohorts cancer groups
all_cohorts_cancer_groups <- current_table %>% 
  dplyr::select(cohort, Disease) %>% 
  dplyr::filter(cohort == "All Cohorts") %>% 
  dplyr::distinct() %>%
  dplyr::rename(current_cohort = cohort) %>% 
  dplyr::full_join(previous_table %>% 
                     dplyr::select(cohort, Disease) %>% 
                     dplyr::filter(cohort == "All Cohorts") %>% 
                     dplyr::distinct() %>% 
                     dplyr::rename(previous_cohort = cohort),
                   by = "Disease") %>% 
  dplyr::select(Disease, current_cohort, previous_cohort)

# write to file
all_cohorts_cancer_groups %>% 
  readr::write_tsv(file.path(results_dir, 
                             "all_cohorts_cancer_groups.tsv"), na = "NA")

# display   
all_cohorts_cancer_groups
```


### Changes in common columns across tpm tables

#### Check gene symbols
```{r gene symbols}
changes_gene_symbols <- 
  changes_in_columns(current_table, previous_table, "Gene_symbol")

# write to file
changes_gene_symbols %>% 
  readr::write_tsv(file.path(results_dir, 
                             "changes_in_gene_symbols.tsv"), na = "NA")

# display   
changes_gene_symbols
```



#### Check Ensembl IDs
```{r ensembl ids}
changes_ensembl_ids <- 
  changes_in_columns(current_table, previous_table, "Gene_Ensembl_ID")

# write to file
changes_ensembl_ids %>% 
  readr::write_tsv(file.path(results_dir, 
                             "changes_in_ensembl_ids.tsv"), na = "NA")

# display   
changes_ensembl_ids
```


#### Check cancer groups
```{r cancer groups}
changes_cancer_groups <- 
  changes_in_columns(current_table, previous_table, "Disease")

# write to file
changes_cancer_groups %>% 
  readr::write_tsv(file.path(results_dir, 
                             "changes_in_cancer_groups.tsv"), na = "NA")

# display   
changes_cancer_groups
```



#### Check EFO IDs
```{r efo ids}
changes_efo_ids <- 
  changes_in_columns(current_table, previous_table, "EFO")

# write to file
changes_efo_ids %>% 
  readr::write_tsv(file.path(results_dir, "changes_in_efo_ids.tsv"), na = "NA")

# display   
changes_efo_ids
```


#### Check MONDO IDs
```{r mondo ids}
changes_mondo_ids <- 
  changes_in_columns(current_table, previous_table, "MONDO")

# write to file
changes_mondo_ids %>% 
  readr::write_tsv(file.path(results_dir, "changes_in_mondo_ids.tsv"), na = "NA")

# display   
changes_mondo_ids
```


### Generate excel output for all QC checks and summaries
```{r}
# read all QC  checks and summaries files 
qc_files <- list.files(results_dir, pattern = "tsv$")
qc_files_list <- lapply(qc_files, function(x) {
  # read each file
  qc_file_df <- readr::read_tsv(file.path(results_dir, x))
  return(qc_file_df)
})
# get file names
qc_files_names <- gsub(".tsv", "", qc_files)
names(qc_files_list) <- qc_files_names
# write results to excel workbook 
qc_files_list %>% 
  openxlsx::write.xlsx(file.path(results_dir, 
                                 paste0(gsub(".tsv.gz", "", 
                                             basename(params$current_table)), 
                                        ".xlsx")), 
                       overwrite = TRUE, keepNA = TRUE, na.string = "NA")
```


```{r session info}
sessionInfo()
```